---
title: "Introduction to curvr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: curvr.bib
csl: cell.csl
link-citations: true #'yes' for html, 'true' for pdfs
linkcolor: blue
urlcolor: blue
citecolor: blue
---


```{r include=FALSE} 
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```

```{r setup}
library(curvr)
library(dplyr)
library(Momocs)
library(pracma)
```

#### Abstract

1. Flower-pollinator curvature as viewed from the side (dorsiventral plane) has been a trait of special interest through the post-Darwin era of pollination ecology. However, within pollination ecology there are at least four metrics in use with few references to their origins or the meaning of the associated units. [@boehm_2020] recently suggested clarifying the defintion of curvature by using concepts of differential geometry. 

2. Here, we introduce an R package ```curvr``` for computing curvature from lines fitted to landmarked specimens. ```curvr``` is designed to take inputs from ```Momocs```, an open-source toolkit for 2D morphometrics in R. We also outline possible advances to the curve-fitting protocol that may subsequently improve estimations of total curvature. 

3. Example dataset. 

4. 

\newpage


#### 1. Introduction

The curvature of flowers and pollinator mouthparts (e.g. bird bills) is a widespread trait contributing to morphological diversity in the angiosperms and their pollinators. The evolutionary ecology and history of measuring curvature in pollination ecology was recently reviewed in [@boehm_2020]. To briefly summarize; the curvature of floral tubes as viewed from the dorsiventral plane (hereafter 'curvature') has long been recognized as morphological feature that contributes to floral isolation [sensu @grant_1949]. As curvature represents one of three spatial dimensions through which floral diversity can evolve (in addition to width and length), it is likely to play a role in floral diversification [@kay_2009; @armbruster_2009; @vamosi_2018]. Similarly, pollinators that specialize on curved flowers evolve curved mouthparts, contributing to morphological diversification of these clades [@weinstein_2017; @maruyama_2018].  Thus, flower-pollinator curvature has been of special interest to pollination ecologists, resulting in over 40 studies since the late 1970's. Among these studies [@boehm_2020] identified four metrics for defining and quantifying curvature, in addition to the more general approach of quantifying shape via geometric morphometrics. Despite the many approaches to measuring curvature there is a lack of consensus as to what curvature means, how to interpret their units, and how varying methods could or should be compared. To this end, [@boehm_2020] proposed a revised definition of curvature that directly borrowed from the mathematical literature, including an interpretation of accompanying units. 

Curvature, as defined in differential geometry, is the rate of change of the tangent, $\mathbf{T}$, with respect to arc length, $s$, along the curve: 

$$\kappa = \frac{d \mathbf{T}}{ds}$$ 

When the angle, $\phi$ of $\mathbf{T}$ is related to an axis of reference (e.g. the $x$ axis), curvature can be expressed as:

$$\kappa = \frac{d\phi}{ds}$$

For a straight line, the direction of the tangent is constant and therefore, curvature is zero. For a curved line, the tangent may drastically change direction from point to point, resulting in some non-zero value of curvature. Integrating point-wise curvature gives the _total curvature_ of a curve:

$$\kappa_{total} = \int_{0}^{s_{max}} \kappa \; ds$$

At some arbitrary point $i$, point-curvature ($\kappa_i$) is reported as $degrees$ per unit length. For example, if measurements have been made at every millimeter along the curve, point-wise curvature is reported as $degrees \cdot mm^{-1}$. _Total curvature_, the sum of total tangent rotations across the length of the curve, is expressed as $degrees$. Further techincal details can be found in [@boehm_2020] and references therein.  

The R package ```curvr``` was created to demonstrate the utility of this revised definition of curvature - it calculates point-wise curvature from curves fitted to landmarked specimens. This package is designed to work alongside common protocols in geometric morphometrics (GM) - the study of shape and its covariance with associated or causal variables [@bookstein_1991; @macleod_2002]. While relevant concepts from GM are outlined here, we recommend the concise introduction to the topic by Webster and Sheets [-@webster_2010], and for detailed reading, Zelditch et al [-@zelditch_2004].  

Geometric morphometrics aims at quantifying shape, i.e. the geometric information remaining after after specimen size, location, and orientation have been accounted for [@kendall_1977; @bookstein_1991]. A typical GM protocol begins with assigning landmarks to the specimen. Landmarks are points on a specimen that are topologically or biologically homologous [@bookstein_1991]. Generally, the more landmarks assigned to a specimen, the more information about its shape that has been captured. In a comparative study, landmarked specimens are scaled, translated, and rotated to remove features not contributing to shape [@rohlf_1990_b]. In _outline-based_ GM, shape functions are fit to the landmarks, for example using the tangent angle function [@zahn_1972]. In _landmark-based_ GM, the set of landmarks are taken together as a 'landmark configuration' and exist in a multidimensional shape space, defined by the number of landmarks used, and the spatial dimensions in which they were assigned [2D or 3D, @kendall_1984]. In both approaches, shape has been summarised as a single variable so that useful statistical procedures can be applied; most GM protocols involve principal components analysis (PCA) of shape variables, enabling the major axes of shape variation to be calculated and plotted in PCA 'shape space' [@macleod_2002]. 

Inspection of deformation grids [@bookstein_1991] or end-member specimens [@macleod_2002] plotted onto these axes are suggestive of the geometric features contributing to shape variation (e.g. curvature). The role of ```curvr``` is to re-purpose the landmark assignments used in the GM analysis to calculate _total curvature_ from each specimen. By explicitly quantifying curvature, the investigator can empircally test the contribution of curvature to shape variation along some principal axis of interest. In the following section the algorithims for computing point-wise curvature are described, followed by a demonstration of its utility alongside a typical GM analysis. 


\newpage


#### 2. The `curvr` package

Computing point-wise and total curvature necessitates fitting a curve to the specimen of interest. Fortunately, the landmarks used in a GM protocol can be re-purposed for regressing curves to the specimen. However, in GM, landmarks are placed throughout the entire specimen to quantify shape. Quantifying curvature typically implies a segment of the shape that is of particular interest [but see: @silk_1978], therefore the landmarks used for curvature analysis will be a subset of those used in the GM protocol. 

The `curvr` package can be used alongside a typical GM protocol where specimens have been digitally landmarked using eg tpsDig [@rohlf_2015], `Momocs` [@bonhomme_2014], or `geomorph` [@adams_2013]. Landmarked specimens with $xy$ coordinates can then be fitted with polynomial curves using `Momocs::npoly.Opn` or `Momocs::opoly.Opn`.  The output of either function is a list-object with the first element containing the polynomial coefficients in ascending order of degrees (note that `Momocs::npoly.default()` creates an `OpnCoe` object - currently, `curvr` is not designed to handle this object class). `curvr::total_curvature` takes the polynomial `list` object(s) as input, extracts the coefficients and constructs an expression that is passed to `stats::deriv3` for computing first and second derivatives. Before point-wise curvature is estimated from the derivatives, the polynomial function is parameterized by arc length. The advantage of this step is that curvature can be calculated at even increments of arc length, $s$. The polynomial is parameterized by arc length by first splitting the curve into sub-curves of incrementally increasing proportions of total length, from 0 to 1. For example, a curve split into 10 parts would have 10 sub-curves with arc length calculated as $0.1(s)$, $0.2(s)$, ..., $1.0(s)$. Each sub-curve has a lower $x$ bound of known value (the lower bound of the original polynomial). The upper $x$ bound (of each sub-surve) is unknown. The root-finding function `stats::uniroot` solves for the unknown upper $x_i$ value, $u_i$, for each $i$ sub-curve:

```f(u) = pracma::arclength(polynomial, lower x boundary, u)$length```

The result is a vector of $x$ values that correspond to even increments of arc length along the polynomial. We then use the gradient (first derivatives, $y'$) and Hessian matricies (second derivatives, $y''$) to calculate point-wise $\kappa_i$ at every position $x_i$:

$$\kappa = \frac{|y''|}{(1+y'^2)^{\frac{3}{2}}}$$ 


Although written differently than the equation for curvature stated in the introduction, this definition is equivalent [@rutter_2000, pp. 133-139]. 

Finally, the point-wise curvature measurements are summed and converted from radians to degrees. The output of `curvr::total_curvature()` is a `list` containing point-wise and  total curvature as defined above. 

Before submitting `Momocs` polynomials to `curvr`, the user first needs to inspect their list-object and extract the lower and upper $x$ bounds of their polynomial (coming soon to Vignettes). These bounds are entered as the `x_range` parameter of `curvr::total_curvature()`. Additionally, the user needs to specify the number of subdivisions made to their curve(s). Greater subdivisions more completely sample the curve, but increases computing time. It is recommended to use subdivisions on the order of 800 or greater. 

In its current form, `curvr` accepts outputs from the R package `Momocs` [@bonhomme_2014], specifically polynomial objects (`list`s, but not `OpnCoe`) fitted to landmarked specimens. Because polynomials will often miss local curvature features, future versions of `curvr` will adopt more sophisticated curve-fitting procedures from the field of outline-based morphometrics (discussed further in final section). 


#### 3. Applications

In this section, we will illustrate the utility of quantifying curvature along a typical analysis of morphometric data. Here, we will test whether bill curvature is correlated with variation along a principal component of shape space. Morphometric has been kindly provided by Navalon et al (2020) - their study examined skull and bill morphology as integrated or decoupled phenotypes. One of the results from Navalon et al (2020, p.17) states: 


-Recreate GM analysis
-complement with curvature analysis


Section: shortcoming of using polynomials. Suggest alternate methods. 
Table 1. Functions available in ```curvr``` version 0.1.0 for computing point-wise curvature

Columns: Function, Method, Reference



In GM analyses, 











We recommend using at least 800 subdivisions:

```
x <- 1:10
y <- x^2
mdat <- matrix(c(x,y), nrow=10, ncol=2)
my_poly <- Momocs::npoly(mdat, 2)
  
K5     <- totalK(c(0,1), my_poly, 5)
K20    <- totalK(c(0,1), my_poly, 20)
K50    <- totalK(c(0,1), my_poly, 50)
K200   <- totalK(c(0,1), my_poly, 200)
K800   <- totalK(c(0,1), my_poly, 800)
K2000  <- totalK(c(0,1), my_poly, 2000)
K10000 <- totalK(c(0,1), my_poly, 10000)

subdiv <- c(5, 20, 50, 200, 800, 2000, 10000)
K_values <- c(55.34299, 46.00986, 44.1399, 43.20422, 42.97015, 42.92333, 42.89837)

sub_vs_K <- matrix(nrow=7, ncol=2)
sub_vs_K[,1] <- subdiv
sub_vs_K[,2] <- K_values

sK_tbl <- as_tibble(sub_vs_K)
ggplot(data=sK_tbl, aes(V1, V2)) + geom_point()


```
example
short: https://besjournals.onlinelibrary.wiley.com/doi/epdf/10.1111/2041-210X.13385 
https://besjournals.onlinelibrary.wiley.com/doi/epdf/10.1111/2041-210X.13293

pgs 184, 186o
https://ebookcentral.proquest.com/lib/ubc/reader.action?docID=422384&ppg=18

#### References 

